---
output: github_document
bibliography: vicar.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Various Ideas for Confounder Adjustment in Regression
[![Windows Build](https://ci.appveyor.com/api/projects/status/github/dcgerard/vicar?branch=master&svg=true)](https://ci.appveyor.com/project/dcgerard/vicar)
[![Linux Build](https://travis-ci.org/dcgerard/vicar.svg?branch=master)](https://travis-ci.org/dcgerard/vicar)
<!-- [![Coverage Status](https://coveralls.io/repos/github/dcgerard/vicar/badge.svg?branch=master)](https://coveralls.io/github/dcgerard/vicar?branch=master) -->
[![Coverage](https://img.shields.io/codecov/c/github/dcgerard/vicar/master.svg)](https://codecov.io/github/dcgerard/vicar?branch=master)

## Description
Let

Y = XB + ZA + E,

for

* Y an n by p matrix of gene expression data with n samples and p genes,
* X an n by q matrix of q covariates,
* B a q by p matrix of unobserved coefficients for the observed covariates,
* Z an n by k matrix of hidden confounders,
* A a k by p matrix of hidden coefficients for the hidden confounders, and
* E an n by p matrix of independent normal errors with column variances s1,...,sp.

Not accounting for the hidden covariates, Z, can reduce power and
result in poor control of false discovery rate. This package provides
a suite of functions to adjust for hidden confounders, both when one
has and does not have access to control genes.

The function `mouthwash` can adjust for hidden confounding when one
does not have access to control genes. It does so via a non-parametric
empirical Bayes method that uses the powerful methodology of Adaptive
SHrinkage [@stephens2016false] within the factor-augmented regression
framework described in @wang2015confounder.

When one has control genes, there are many approaches to take. Such
methods include RUV2 [@gagnon2012using], RUV4 [@gagnon2013removing],
and CATE [@wang2015confounder]. This package adds to the field of
confounder adjustment with control genes by

1. Implementing a version of CATE that is calibrated using control genes similarly to the method in @gagnon2013removing. The function is called `vruv4`.
2. Introduces RUV3, a version of RUV that can be considered both RUV2 and RUV4. The function is called `ruv3`.
3. Introduces RUV-impute, a more general framework for accounting for hidden confounders in regression. The function is called `ruvimpute`
4. Introduces RUV-Bayes, a Bayesian version of RUV. The function is called `ruvb`.

Please cite this package as:

> Gerard, David. 2016. *vicar: Methods for Confounder Adjustment in Regression*. [https://github.com/dcgerard/vicar](https://github.com/dcgerard/vicar).

Or, using BibTex:
``` tex
@Manual{gerard2016vicar,
    title = {{vicar}: Various Ideas for Confounder Adjustment in Regression},
    author = {David Gerard},
    year = {2016},
    note = {R package version 0.1.5},
    url = {https://github.com/dcgerard/vicar},
}
```

See also the related R packages
[`cate`](https://cran.r-project.org/web/packages/cate/index.html)
[@wang2015cate] and
[`ruv`](https://cran.r-project.org/web/packages/ruv/index.html)
[@bartsch2015ruv].

Check out [NEWS.md](NEWS.md) to see what's new with each version.

## Installation
To install, run in R:

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("dcgerard/vicar")
```

## References
