---
output: github_document
bibliography: vicar.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# VICAR: Variance Inflation for Confounder Adjustment in Regression
[![Build Status](https://travis-ci.org/dcgerard/vicar.svg?branch=master)](https://travis-ci.org/dcgerard/vicar)
[![Coverage Status](https://coveralls.io/repos/github/dcgerard/vicar/badge.svg?branch=master)](https://coveralls.io/github/dcgerard/vicar?branch=master)

## Description
Let

Y = XB + ZA + E,

for

* Y an n by p matrix of gene expression data with n samples and p genes,
* X an n by q matrix of of q covariates,
* B a q by p matrix of unobserved coefficients for the observed covariates,
* Z an n by k matrix of hidden confounders,
* A a k by p matrix of hidden coefficients for the hidden confounders, and
* E an n by p matrix of independent normal errors with column variances s1,...,sp.

Not accounting for the hidden covariates, Z, can reduce power and result in poor control of false discovery rate.

There are many approaches available for estimating hidden
confounders. This package is among those that use control genes to
account to hidden confounding.  Other such methods include RUV2
[@gagnon2012using], RUV4 [@gagnon2013removing], and CATE
[@wang2015confounder]. This package adds to this field by

1. Implementing a version of CATE that is calibrated using control genes similarly to the method in @gagnon2013removing. The function is called `vruv4`.
2. Introduces RUV3, a version of RUV that can be considered both RUV2 and RUV4. The function is called `ruv3`.
3. Introduces RUV-impute, a more general framework for accounting for hidden confounders in regression. The function is called `ruvimpute`
4. Introduces RUV-Bayes, a Bayesian version of RUV. The function is called `ruvb`.

Please cite this package as:

> Gerard, David. 2016. *vicar: Variance Inflation for Confounder Adjustment in Regression*. [https://github.com/dcgerard/vicar](https://github.com/dcgerard/vicar).

Or, using BibTex:
``` tex
@Manual{gerard2016vicar,
    title = {{vicar}: Variance Inflation for Confounder Adjustment in Regression},
    author = {David Gerard},
    year = {2016},
    note = {R package version 0.1.2},
    url = {https://github.com/dcgerard/vicar},
}
```

See also the related R packages
[`cate`](https://cran.r-project.org/web/packages/cate/index.html)
[@wang2015cate] and
[`ruv`](https://cran.r-project.org/web/packages/ruv/index.html)
[@bartsch2015ruv].

Check out [NEWS.md](NEWS.md) to see what's new with each version.

## Installation
To install, run in R:

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("dcgerard/vicar")
```

## References
