---
output: github_document
bibliography: vicar.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# VICAR: Variance Inflation for Confounder Adjustment in Regression
[![Build Status](https://travis-ci.org/dcgerard/vicar.svg?branch=master)](https://travis-ci.org/dcgerard/vicar)

## Description
Let

Y = XB + ZA + E,

for

* Y an n by p matrix of gene expression data with n samples and p genes,
* X an n by q matrix of of q covariates,
* B a q by p matrix of unobserved coefficients for the observed covariates,
* Z an n by k matrix of hidden confounders,
* A a k by p matrix of hidden coefficients for the hidden confounders, and
* E an n by p matrix of independent normal errors with column variances s1,...,sp.

Not accounting for the hidden covariates, Z, can reduce power and result in poor control of false discovery rate.

There are many approaches available for estimating hidden confounders,
but in practice they are not well calibrated. This package calibrates
RUV2 [@gagnon2012using], RUV4 [@gagnon2013removing], and LEAPP
[@sun2012multiple]. If I come up with a way to calibrate SVA
[@leek2007capturing, @leek2008general], then I'll include that as
well.

Check out [NEWS.md](NEWS.md) to see what's new with each version.

## Installation
To install, run in R:

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("dcgerard/vicar")
```

## References
