---
title: "Various Ideas for Confounder Adjustment in Regression"
output: github_document
bibliography: "./vignettes/vicar.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Windows Build](https://ci.appveyor.com/api/projects/status/github/dcgerard/vicar?branch=master&svg=true)](https://ci.appveyor.com/project/dcgerard/vicar)
[![Linux Build](https://travis-ci.org/dcgerard/vicar.svg?branch=master)](https://travis-ci.org/dcgerard/vicar)
[![Coverage Status](https://coveralls.io/repos/github/dcgerard/vicar/badge.svg?branch=master)](https://coveralls.io/github/dcgerard/vicar?branch=master)
[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)


## Description
Let

Y = XB + ZA + E,

for

* Y an n by p matrix of gene expression data with n samples and p genes,
* X an n by q matrix of q covariates,
* B a q by p matrix of unobserved coefficients for the observed covariates,
* Z an n by k matrix of hidden confounders,
* A a k by p matrix of hidden coefficients for the hidden confounders, and
* E an n by p matrix of independent normal errors with column variances s1,...,sp.

Not accounting for the hidden covariates, Z, can reduce power and
result in poor control of false discovery rate. This package provides
a suite of functions to adjust for hidden confounders, both when one
has and does not have access to control genes.

The functions `mouthwash()` and `backwash()` can adjust for hidden confounding when one
does not have access to control genes. They do so via non-parametric
empirical Bayes methods that use the powerful methodology of Adaptive
Shrinkage [@stephens2016false] within the factor-augmented regression
framework described in @wang2017confounder. `backwash()` is a slightly more 
Bayesian version of `mouthwash()`. These methods are described in
@gerard2018empirical.

When one has control genes, there are many approaches to take. Such
methods include RUV2 [@gagnon2012using], RUV4 [@gagnon2013removing],
and CATE [@wang2017confounder]. This package adds to the field of
confounder adjustment with control genes by

1. Implementing a version of CATE that is calibrated using control genes similarly to the method in @gagnon2013removing. The function is called `vruv4()`.
2. Introduces RUV3, a version of RUV that can be considered both RUV2 and RUV4. The function is called `ruv3()`.
3. Introduces RUV-impute, a more general framework for accounting for hidden confounders in regression. The function is called `ruvimpute()`
4. Introduces RUV-Bayes, a Bayesian version of RUV. The function is called `ruvb()`.

These additions are described in detail in @gerard2019unifying.

See also the related R packages
[`cate`](https://cran.r-project.org/package=cate)
[@wang2015cate] and
[`ruv`](https://cran.r-project.org/package=ruv)
[@bartsch2015ruv].

Check out [NEWS.md](NEWS.md) to see what's new with each version.

## How to cite

If you use any of the control-gene based methods, please cite:

> Gerard, D., & Stephens, M. 2019. "Unifying and Generalizing Methods for Removing Unwanted Variation Based on Negative Controls". *Statistica Sinica*, in press <[doi:10.5705/ss.202018.0345](https://doi.org/10.5705/ss.202018.0345)>.

Or, using BibTex:
``` tex
@article{gerard2019unifying,
  title={Unifying and Generalizing Methods for Removing Unwanted Variation Based on Negative Controls},
  author={Gerard, David and Stephens, Matthew},
  journal={Statistica Sinica},
  doi={10.5705/ss.202018.0345},
  volume={},
  number={},
  year={2019},
  pages={in press}
}
```

If you use either MOUTHWASH or BACKWASH, please cite:

> Gerard, D., & Stephens, M. 2018. "Empirical Bayes shrinkage and false discovery rate estimation, allowing for unwanted variation", *Biostatistics*, , kxy029, <[doi:10.1093/biostatistics/kxy029](https://doi.org/10.1093/biostatistics/kxy029)>.

Or, using BibTex:
``` tex
@article{gerard2018empirical,
  author = {Gerard, David and Stephens, Matthew},
  title = {Empirical {B}ayes shrinkage and false discovery rate estimation, allowing for unwanted variation},
  journal = {Biostatistics},
  volume = {},
  number = {},
  pages = {kxy029},
  year = {2018},
  doi = {10.1093/biostatistics/kxy029},
}
```

Otherwise, please cite this package as:

> Gerard, David. 2016. *vicar: Various Ideas for Confounder Adjustment in Regression*. [https://github.com/dcgerard/vicar](https://github.com/dcgerard/vicar).

Or, using BibTex:
``` tex
@Manual{gerard2016vicar,
    title = {{vicar}: Various Ideas for Confounder Adjustment in Regression},
    author = {David Gerard},
    year = {2016},
    note = {R package version 0.1-9},
    url = {https://github.com/dcgerard/vicar},
}
```

## Installation

To install, first install `sva` and `limma` from Bioconductor
in R:

```{r, eval = FALSE}
install.packages("BiocManager")
BiocManager::install(c("limma", "sva"))
```

Then run in R:

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("dcgerard/vicar")
```

If you want some of the tools in `vicar` to be exactly equivalent to those in `ruv`, you'll need to install an older version of `ruv` (`ruv` was updated and now the those equivalencies are not *exactly* the same)

```{r, eval=FALSE}
devtools::install_version("ruv", version = "0.9.6", repos = "http://cran.us.r-project.org")
```

A note about matrix computations in vicar: Some of the methods in the vicar package such as mouthwash and backwash rely heavily on matrix-vector operations. The speed of these operations can have a big impact on vicar's performance, especially in large-scale data sets. If you are applying vicar to large data sets, I recommend that you set up R with optimized BLAS (optionally, LAPACK) libraries, especially if you have a multicore computer (most modern laptops and desktops are multicore). See [here](https://csgillespie.github.io/efficientR/set-up.html#blas-and-alternative-r-interpreters) and [here](https://cran.r-project.org/doc/manuals/r-release/R-admin.html#Linear-algebra) for advice and technical details on this. For example, in [our experiments on a high-performance compute cluster](https://github.com/pcarbo/mouthwash_sims/blob/master/mouthwash.sbatch) we set up R with multithreaded OpenBLAS.


## Vignettes
I've provided three vignettes to help you get started with vicar. By default, the vignettes are not built when you use `install_github`. To build the vignettes during installation, run
```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("dcgerard/vicar", build_vignettes = TRUE)
```
Note that this will result in a somewhat slower install. The first vignette, _sample\_analysis_, gives a sample analysis using vicar to account for hidden confounding. The second vignette, _customFA_, gives a few instructions on how to incorporate user-defined factor analyses with the confounder adjustment procedures implemented in vicar. The third vignette, _custom\_prior_, gives instructions and examples on incorporating a user-specified prior into `ruvb`. To see these vignettes after install, run
```{r, eval = FALSE}
utils::vignette("sample_analysis", package = "vicar")
utils::vignette("customFA", package = "vicar")
utils::vignette("custom_prior", package = "vicar")
```

## References
