---
title: "Custom Prior for RUVB"
author: "David Gerard"
date: "2016-12-13"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Prior for RUVB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Abstract

I provide instructions on how to incorporate a user-specified prior into `ruvb`. 
I provide an example with a simulated RNA-seq dataset.

# Data

We will use the simulated RNA-seq dataset described in the vignette _sample\_analysis_. To read a description about these data, run the following in R:
```{r, eval = FALSE}
utils::vignette("sample_analysis", package = "vicar")
```

We'll first load in these data:
```{r, results='hide'}
library(vicar)
library(ggplot2)
set.seed(2512)
data(sim_gtex)
Y <- sim_gtex$Y
X <- sim_gtex$X
ctl <- sim_gtex$ctl
which_null <- sim_gtex$which_null
beta <- sim_gtex$beta
```

As in _sample\_analysis_, we'll estimate the number of hidden confounders using the `num.sv` function from the sva R package.
```{r}
num_sv <- sva::num.sv(dat = t(Y), mod = X, method = "be")
num_sv
```

# Specifying a Prior

In the `ruvb` function, one can use the `prior_fun` and `prior_args` parameters to specify a prior in the second step of RUVB. If $p$ is the number of genes, $m$ is the number of control genes, and $k$ is the number of covariates of interest, then `prior_fun` takes the following as input

1. `beta_mat`: An $k$ by $p - m$ matrix.
2. Anything else the user desires, specified in the list `prior_args`.

By specifying the `return_log` argument in `ruvb`, `prior_fun` can either return the density (a non-negative numeric scalar), or the log-density (a real numeric scalar). For numerical stability reasons, generally you should have `prior_fun` return the log-density and then set `return_log = TRUE`.

# Example

Suppose we had slightly stronger prior information that the signals are all relatively weak, so we'll put a $N(0, 5)$ prior on the effects. For numerical stability, we'll return the log-density.

```{r}
strong_prior <- function(beta_mat, sd_prior) {
  sum(stats::dnorm(beta_mat, mean = 0, sd = sd_prior, log = TRUE))
}
```

Since `sd_prior` is an argument we need to specify, we'll use the `prior_args` parameter to set `sd_prior = 5`. Now let us run `ruvb`. Note that you should run the MCMC for many more iterations than what I'm doing here.
```{r}
ruvbout <- ruvb(Y = Y, X = X, ctl = ctl, k = num_sv, cov_of_interest = 2,
                prior_fun = strong_prior, prior_args = list(sd_prior = 5),
                return_log = TRUE,
                fa_args = list(display_progress = FALSE, nsamp = 1000, thin = 1))
```

We'll compare the results against just using the default uniform prior:
```{r}
ruvbout_unif <- ruvb(Y = Y, X = X, ctl = ctl, k = num_sv, cov_of_interest = 2,
                     fa_args = list(display_progress = FALSE, nsamp = 1000, thin = 1))
```

They give about the same AUC's.
```{r}
aucnorm <- pROC::roc(response = which_null[!ctl], predictor = c(ruvbout$lfsr2))$auc
aucunif <- pROC::roc(response = which_null[!ctl], predictor = c(ruvbout_unif$lfsr2))$auc
aucdat <- data.frame(Prior = c("Normal", "Uniform"), AUC = c(aucnorm, aucunif))
knitr::kable(x = aucdat)
```
